tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-20 paper.tex
Minor updates

An important property of the succinct tree sequence ARG encoding
used by these population genetic simulators is that it allows us to
store enormous genetic variation datasets compactly. Instead of
storing the genetic sequence variation in the conventional
matrix-based form~\citep[e.g.\ VCF,][]{danecek2011},
we can store the underlying genealogical history
and mutations that ultimately \emph{result} in the observed
variation~\citep{kelleher2019}.
For example, a highly realistic simulation
of chromosome 9 for 1.4 million
French-Canadian individuals~\citep{anderson2023}
requires around 550GB of storage space in
gzip-compressed VCF ($\sim$280TB uncompressed).
The original simulation file representing
precisely the same genetic variation data, plus the
complete simulated ARG, requires only 1.36GB of space
in compressed \texttt{tszip} format. Thus,
by encoding in terms of the underlying ARG,
in this case we achieve more than 400-fold compression,
and this highly unwieldy dataset
becomes straightforward to work with (we discuss this
landmark dataset again in later sections).
Another substantial benefit of storing simulated datasets
in this ARG-based format is that many computations can
be expressed efficiently in terms of the underlying
ancestry~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.

There are therefore sophisticated methods available for simulating
ARGs at the population scale, but there is currently no
easy way to simulate quantitative traits based on such an ARG.
Existing methods to simulate quantitative traits from a given set of
genetic sequences in standard formats~\citep[e.g.][]{meyer2018,fernandes2020}
are overwhelmed by the scale of the data. For example,
just parsing the 280TB of VCF for the French-Canadian
dataset mentioned above would take many days of CPU time,
making quantitative trait simulation by this approach unfeasible.
% TODO this needs another pass. We do need to address the Martin et
% al thing up front though, I think.
In their analysis of the portability of polygenic risk
scores across populations, \citet{martin2017}
used efficient \texttt{tskit} APIs to simulate quantitative
traits for 600,000 humans directly from a simulated ARG.
While the code provided by Martin et al.\ works at this
very large scale, it is tightly coupled to the details of their study
and not designed to be reused.
Simulation code can be subtle and difficult to
debug~\citep{ragsdale2020lessons}, and there
is a critical need for a well-documented and
thoroughly tested means of simulating quantitative
traits directly from an ARG.

@@

\section{Conclusion}
We believe that in the coming years, simulation in tree sequence encoding will
attract more users in need of conducting genetic simulations with complex
population structure. Multiple studies highlight the advantages of using the
ARG data structure in GWAS \citep{link2023tree,nowbandegani2023extremely,zhang2023}.
ARG data
structure is becoming increasingly useful in genetic studies, as it is possible
to accurately infer biobank-scale genealogies from sequencing data
\citep{zhang2023}, and simulate large sample whole-genome sequences with
spatiotemporal metadata \citep{anderson2023}. These advances in ARG presents a
real opportunity for people to further understand the effects of complex
population structure on quantitative traits. The \texttt{tstrait} package is an
essential piece of infrastructure that lets people explore these possibilities.

>>>>

An important property of the succinct tree sequence ARG encoding
used by these population genetic simulators is that it allows us to
store enormous genetic variation datasets compactly. Instead of
storing the genetic sequence variation in the conventional
matrix-based form~\citep[e.g.\ VCF,][]{danecek2011},
we can store the underlying genealogical history
and mutations that ultimately \emph{result} in the observed
variation~\citep{kelleher2019}.
For example, a highly realistic simulation
of chromosome 9 for 1.4 million
French-Canadian individuals~\citep{anderson2023}
requires around 550GB of storage space in
gzip-compressed VCF ($\sim$280TB uncompressed).
The original simulation file representing
precisely the same genetic variation data, plus the
complete simulated ARG, requires only 1.36GB of space
in compressed \texttt{tszip} format. Thus,
by encoding in terms of the underlying ARG,
in this case we achieve more than 400-fold compression,
and this highly unwieldy dataset
becomes straightforward to work with (we discuss this
landmark dataset again in later sections).
Another substantial benefit of storing simulated datasets
in this ARG-based format is that many computations can
be expressed efficiently in terms of the underlying
genealogies~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.

Although there are sophisticated methods available for simulating
ARGs at the population scale, there is currently no
easy way to simulate quantitative traits based on such an ARG.
Existing methods to simulate quantitative traits from a given set of
genetic sequences in standard formats~\citep[e.g.][]{meyer2018,fernandes2020}
are overwhelmed by the scale of the data. For example,
just parsing the 280TB of VCF for the French-Canadian
dataset mentioned above would take many days of CPU time,
making quantitative trait simulation by this approach unfeasible.
In their analysis of the portability of polygenic risk
scores across populations, \citet{martin2017}
demonstrated the ability to simulate phenotypes at scale using ARGs. 
However, their approach is tightly coupled to the details 
of the study 
and not designed to be reused.
Simulation code can be subtle and difficult to
debug~\citep{ragsdale2020lessons}, and there
is a critical need for a well-documented and
thoroughly tested means of simulating quantitative
traits directly from an ARG.

@@

\section{Conclusion}
We believe that in the coming years, simulation in tree sequence encoding will
attract more users in need of conducting genetic simulations with complex
population structure. Multiple studies highlight the advantages of using the
ARG data structure in GWAS \citep{link2023tree,nowbandegani2023extremely,zhang2023}.
ARG data
structures are becoming increasingly useful in genetic studies, as it is possible
to accurately infer biobank-scale genealogies from sequencing data
\citep{zhang2023}, and simulate large sample whole-genome sequences with
spatiotemporal metadata \citep{anderson2023}. These advances in ARG presents a
real opportunity for people to further understand the effects of complex
population structure on quantitative traits. The \texttt{tstrait} package is an
essential piece of infrastructure that lets people explore these possibilities.
